\texttt{disttree}: Distributional Trees and Forests

### 2017-09-09 pkg/disttree/inst/draft/draft.Rnw
added optional argument 'noise_sd' to set the standard deviation of the noise variable within the data generating process 'dgp()'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1567 255f4249-2263-4417-b1ea-8498779e27a7

######## data generating process based on a predefined tree or a parameter function
# if tree: 2 splits => 3 terminal nodes
# 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
#
# input: n            ..... nr of observations
#        family       ..... distribution of the generated observations
#        split.matrix ..... indices of the variables used for the splits together with the corresponding split points
#        parm.matrix  ..... set of distribution parameters for each subgroup
#        fun          ..... parameter function
#
# output: data.frame with generated observations y (generated seperatly in each subgroups with given distributions parameters),
#                         the given split variables x1, ..., x10 for each observation and
#                         the index of the subgroup for each observation or
#                         the distribution parameter for each observation

# FIX: until now only complete lists can be handed over (dist_list...)
dgp <- function(n, family = dist_list_normal, 
                fun = NULL,
                split.matrix = matrix(nrow = 2, ncol = 2), 
                par.matrix = matrix(nrow = 3, ncol = 2), 
                round.sp = 3)
{

@@

@@

### plotting function for one data set
plot1 <- function(learndata, dt=NULL, df=NULL, g=NULL, all.par = FALSE, df.par = FALSE, g.par = FALSE, dt.par = FALSE, all.var = FALSE) 
{
  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
                    df$fitted.par,
                    dt$fitted.par,
                    #coef(dt)[paste(dt$fitted[,1]),],
                    g$mu.fv, g$sigma.fv)

@@

In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables. This idea was extended to generalized linear models (GLM) which broadened the range of possible distributions of the response variable to the exponential family. (In R these models can be built using the functions \code{lm()} and \code{glm()} provided by the package \pkg{stats}).\\
Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination. (In the fitting function \code{gam} types of smoothing terms can be selected.)\\
However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
(GAMLSS) (\cite{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM. In that way a whole distribution can be specified. The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
This development in the field of parametric modeling is illustrated in figure \ref{Devel_parmod}.\\

@@

All these parametric models work very well as long as only additive effects appear in the considered data. A different type of model that can capture non-additive effects as well are regression trees. Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node. One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of figure \ref{Devel_treeforest}. One way how to tackle this problem is to not just consider one tree but an ensemble of trees. This is the idea of random forests. A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set. Then the results of all trees are combined to get a final model which unlike a tree model can also deal with smooth effects in a reasonable way.\\ 

@@

These (future) steps are illustrated in figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\

@@

In figure \ref{plot_mcycle} the results of a distributional forest are plotted on the top left and then compared to the results of a distributional tree, a GAMLSS and a random forest, regarding the location parameter on the top right and the scale parameter below.\\

@@

\newpage
\section{Application}
\subsection{Continuation of the mcycle example: Cross Validation}
In the introduction the four functions \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest} have been applied on the mcycle data set and their results haven been visualized and discussed. To get a more profound impression of their performances we now want to compare them in a cross validation framework on this text book example data set. The 133 observations are randomly split into 10 subsets roughly of the same size. After learning the model on 9 of these subgroups the RMSE of the estimated location parameter and the log-likelihood function are evaluated on the 10th part. Each subset is chosen once as a test data set and the mean of the resulting values is stored. This procedure is repeated 9 times. This comparison is illustrated in figure \ref{plot_cross_mean}, based on the RMSE on the left and based on the log-likelihood on the right. (Extrapolation of the GAMLSS would lead to very small values of the log-likelihood. To avoid this, the first and the last observation are always included in the learn data set.) \\
As we have already seen in the introduction the distributional tree obviously struggles with the smooth effects while the two forest models can capture the changes quite well. As far as the RMSE of the estimated location parameter is concerned the GAMLSS plays to its strength as a smooth parametric model and leads to the best results followed by the distributional forest and the random forest which might both be a bit to wiggly.\\
However, regarding the log-likelihood the forests perform better than the GAMLSS. This is mainly due to the fact that the GAMLSS models the scale parameter close to zero for observations with \code{times} $<10$. This leads to very low values of the log-likelihood function if one or more of these observations are included in the test data set. As the estimated scale parameter of the tree based models are higher for these specific observations this problem does not occure here.\\
Overall it can be seen that the distributional forest which fits the whole distribution and considers splits in both parameters performs better than the random forest which only models the mean and therefore also only inlcudes splits in the location parameter.

@@

<<echo=FALSE, results=hide>>=
#### cross validation
if(file.exists("simres.rda")){
  load("simres.rda")
} else {
  #source("~/svn/partykit/pkg/disttree/paper/gensim.R")
  source("gensim.R")
  simres <- gensim(seedconst = 7, nrep = 20, ntree = 100L,
                   cov.sep = TRUE,
                   fix.mu = FALSE,
                   fix.sigma = FALSE,
                   mu.sigma.interaction = FALSE,
                   gamboost_cvr = FALSE,
                   eval_disttree = TRUE,
                   eval_distforest = TRUE,
                   eval_gamlss = TRUE,
                   eval_bamlss = TRUE,
                   eval_gamboostLSS = TRUE,
                   eval_randomForest = TRUE,
                   eval_cforest = TRUE)
  save(simres, file = "simres.rda")
}
@

>>>>

######## data generating process based on a predefined tree or a parameter function
  # if tree: 2 splits => 3 terminal nodes
  # 10 split variables (x1, ..., x10) are availabe (5 numeric, 2 binary, 3 categorical)
  #
  # input: n            ..... nr of observations
  #        family       ..... distribution of the generated observations
  #        split.matrix ..... indices of the variables used for the splits together with the corresponding split points
  #        parm.matrix  ..... set of distribution parameters for each subgroup
  #        fun          ..... parameter function
  #        noise_sd     ..... optional parameter to set the standard deviation of the noise variable
  #
  # output: data.frame with generated observations y (generated seperatly in each subgroups with given distributions parameters),
  #                         the given split variables x1, ..., x10 for each observation and
  #                         the index of the subgroup for each observation or
  #                         the distribution parameter for each observation

  # FIX: until now only complete lists can be handed over (dist_list...)
  dgp <- function(n, family = dist_list_normal, 
                  fun = NULL,
                  split.matrix = matrix(nrow = 2, ncol = 2), 
                  par.matrix = matrix(nrow = 3, ncol = 2), 
                  round.sp = 3,
                  noise_sd = 0)
  {

@@

    if(noise_sd > 0) d$y <- d$y + rnorm(n, mean = 0, sd = noise_sd)

@@

### plotting function for one data set
plot1 <- function(learndata, 
                  dt=NULL, df=NULL, g=NULL, gb=NULL, b=NULL, rf=NULL, cf=NULL, 
                  mu=TRUE, sigma=TRUE) 
{
  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
                    df$fitted.par,
                    dt$fitted.par,
                    #coef(dt)[paste(dt$fitted[,1]),],
                    g$mu.fv, g$sigma.fv)

@@

In the field of regression analysis many statistical models have already been developed to model the relationship between a response variable and one or more explanatory variables.\\
Among the first were linear models (LM) which predict the (conditional) expected value of the response variable as a linear combination of the explanatory variables. This idea was extended to generalized linear models (GLM) which broadened the range of possible distributions of the response variable to the exponential family. (In R these models can be built using the functions \code{lm()} and \code{glm()} provided by the package \pkg{stats}).\\
Then, in order to also allow for non-linear effects, generalized additive models (GAM) were introduced where smooth functions of the explanatory variables are summed up instead of the linear combination. (In the fitting function \code{gam} types of smoothing terms can be selected.)\\
However, with these methods only the mean of the distribution of the response variable can be modeled. Generalized additive models for location, scale and shape
(GAMLSS) (\cite{stasinopoulos2005}) made it possible to model each of the parameters of a distribution separately by its own GAM. In that way a whole distribution can be specified. The R-packages \pkg{gamlss} and \pkg{gamlss.dist} offer the corresponding software with a wide variety of distributions.\\
This development in the field of parametric modeling is illustrated in Figure \ref{Devel_parmod}.\\

@@

All these parametric models work very well as long as only additive effects appear in the considered data. A different type of model that can capture non-additive effects as well are regression trees. Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node. One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of Figure \ref{Devel_treeforest}. One way how to tackle this problem is to not just consider one tree but an ensemble of trees. This is the idea of random forests. A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set. Then the results of all trees are combined to get a final model which unlike a tree model can also deal with smooth effects in a reasonable way.\\ 

@@

These (future) steps are illustrated in Figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\

@@

In Figure \ref{plot_mcycle} the results of a distributional forest are plotted on the top left and then compared to the results of a distributional tree, a GAMLSS and a random forest, regarding the location parameter on the top right and the scale parameter below.\\

@@

\newpage
\section{Application}
\subsection{Continuation of the mcycle example: cross validation}
In the introduction the four functions \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest} have been applied on the mcycle data set and their results haven been visualized and discussed. To get a more profound impression of their performances we now want to compare them in a cross validation framework on this text book example data set. The 133 observations are randomly split into 10 subsets roughly of the same size. After learning the model on 9 of these subgroups the RMSE of the estimated location parameter and the log-likelihood function are evaluated on the 10th part. Each subset is chosen once as a test data set and the mean of the resulting values is stored. This procedure is repeated 9 times. This comparison is illustrated in Figure \ref{plot_cross_mean}, based on the RMSE on the left and based on the log-likelihood on the right. (Extrapolation of the GAMLSS would lead to very small values of the log-likelihood. To avoid this, the first and the last observation are always included in the learn data set.) \\
As we have already seen in the introduction the distributional tree obviously struggles with the smooth effects while the two forest models can capture the changes quite well. As far as the RMSE of the estimated location parameter is concerned the GAMLSS plays to its strength as a smooth parametric model and leads to the best results followed by the distributional forest and the random forest which might both be a bit to wiggly.\\
However, regarding the log-likelihood the forests perform better than the GAMLSS. This is mainly due to the fact that the GAMLSS models the scale parameter close to zero for observations with \code{times} $<10$. This leads to very low values of the log-likelihood function if one or more of these observations are included in the test data set. As the estimated scale parameter of the tree based models are higher for these specific observations this problem does not occure here.\\
Overall it can be seen that the distributional forest which fits the whole distribution and considers splits in both parameters performs better than the random forest which only models the mean and therefore also only inlcudes splits in the location parameter.

@@

<<echo=FALSE, results=hide>>=
#### cross validation
if(file.exists("simres.rda")){
  load("simres.rda")
} else {
  #source("~/svn/partykit/pkg/disttree/paper/gensim.R")
  source("gensim.R")
  simres <- gensim(seedconst = 7423, nrep = 20, ntree = 100L,
                   cov.sep = TRUE,
                   fix.mu = FALSE,
                   fix.sigma = FALSE,
                   mu.sigma.interaction = FALSE,
                   gamboost_cvr = FALSE,
                   eval_disttree = TRUE,
                   eval_distforest = TRUE,
                   eval_gamlss = TRUE,
                   eval_bamlss = TRUE,
                   eval_gamboostLSS = TRUE,
                   eval_randomForest = TRUE,
                   eval_cforest = TRUE)
  save(simres, file = "simres.rda")
}
@
